Performance Models for Data Transfers: A Case Study with Molecular Chemistry Kernels

### 2019-04-11 ICPP-19/communications.tex
clean-up

	%%	\todo[inline]{SKumar: Modify both proofs }
	\section{Data Transfer Problem Formulation}
	\label{sec:theoreticalProof}

@@

	%%	\todo[inline]{Mention that memory occupation of a task is the same as its communication time. Skumar: it is not required at all, but we use this for simplicity in Theorem~\ref{th:npComplete}}
	%%	\todo[inline]{Mention that it is always best to use the same order on both machines.}

@@

	\begin{table}
		\begin{center}
			\begin{tabular}{|c|c|c|c|}
				\hline
				%%				\multirow{2}{*}{Task} & Comm Time & \multirow{2}{*}{Comp Time}\\ 
				%%				&=Comm Volume& \\ \hline
				Task & Comm Volume & Comm Time & Comp Time\\ \hline 
				A & 0 & 0 & 5\\ \hline
				B & 4 & 4 & 3\\ \hline
				C & 1 & 1 & 6\\ \hline
				D & 3 & 3 & 7\\ \hline
				E & 6 & 6 & 0.5\\ \hline
				F & 7 & 7 & 0.5\\ \hline
			\end{tabular}
		\end{center}
		\caption{Example instance where ordering on both resources
			has to be different.}
		\label{table:different.order}
	\end{table}

@@

	%%%%name = "optimal_time_infinite_case";
	%%%%string alg_name="order_of_optimal_strategy_infinite_memory";
	%%%%string alg_name="increasing_order_of_communication_strategy";
	%%%%string alg_name="decreasing_order_of_computation_strategy";
	%%%%string alg_name="increasing_order_of_communication_plus_computation_strategy";
	%%%%string alg_name="decreasing_order_of_communication_plus_computation_strategy";
	%%%%string alg_name="optimal_order_infinite_memory_largest_communication\_task\_respects\_memory\_restriction";
	%%%%string alg_name="optimal_order_infinite_memory_smallest_communication\_task\_respects\_memory\_restriction";
	%%%%string alg_name="optimal_order_infinite_memory_maximum_accelerated\_task\_respects\_memory\_restriction";

@@

	\subsection{Dynamic Selection}
	%%%%string alg_name="largest_communication \_ \allowbreak task \_ \allowbreak respects \_ \allowbreak memory \_ \allowbreak restriction";
	%%%%string alg_name="smallest_communication \_ \allowbreak task \_ \allowbreak respects \_ \allowbreak memory \_ \allowbreak restriction";
	%%%%string alg_name="maximum_accelerated \_ \allowbreak task \_ \allowbreak respects \_ \allowbreak memory \_ \allowbreak restriction";

	Dynamic strategies are based on the following principle: when the communication resource is idle, a task is chosen based on a ranking criterion which differ depending on the heuristic, among those which fit in memory. For example, if the selection criteria is to choose a highly compute intensive task, then we compute the ratio of compute time and communication time for all tasks, and we select the task which fit in the currently available memory and whose ratio is maximum (ties are broken arbitrarily). If no task satisfies the selection criteria thee we leave the resource idle at that point and proceed to the next event point. We also ensure that the order on both resources is the same, by ordering tasks on the computing resource accordingly.

@@

	%% We select a set of tasks which induce minimum idle time on compute resource and satisfy available memory requirement. Then a task is chosen for processing from this set based on the following heuristics.
	\begin{enumerate}[label=\roman*)]
		%%	\item $largest_communication \_ \allowbreak task \_ \allowbreak respects \_ \allowbreak memory \_ \allowbreak restriction$: We select a set of tasks which induce minimum idle time on compute channel. A task with the largest communication time is chosen from this set. Memory requirement of this task should not be more than presently available memory. 
		%%	\item $smallest_communication \_ \allowbreak task \_ \allowbreak respects \_ \allowbreak memory \_ \allowbreak restriction$: A task with the smallest communication time is chosen. Memory requirement of this task should not be more than presently available memory. 
		%%	\item $maximum_accelerated \_ \allowbreak task \_ \allowbreak respects \_ \allowbreak memory \_ \allowbreak restriction$: We select a set of tasks which induce minimum idle time on compute channel. A task with the maximum ratio of compute time to communication time is chosen from this set. Memory requirement of this task should not be more than presently available memory.

		\item \textit{largest communication task respects memory restriction} ($LCMR$): a task with the largest communication time is chosen. 
		\item \textit{smallest communication task respects memory restriction} ($SCMR$): a task with the smallest communication time is chosen.
		\item \textit{maximum accelerated task respects memory restriction} ($MAMR$): a task with the maximum ratio of compute time to communication time is chosen.
	\end{enumerate}
	\begin{table}[htb]
		\begin{center}

			\begin{tabular}{|c|c|c|c|}
				\hline
				%%				\multirow{2}{*}{Task} & Comm Time & \multirow{2}{*}{Comp Time}\\ 
				%%				&=Comm Volume& \\ \hline
				Task & Comm Volume & Comm Time & Comp Time\\ \hline 
				A & 3 & 3 & 2\\ \hline
				B & 1 & 1 &  6\\ \hline
				C & 4 & 4 & 6\\ \hline
				D & 5 & 5 & 1\\ \hline
			\end{tabular}
			\caption{\label{tab:dynamicSelectionExample} A task set for dynamic schedules.}
		\end{center}
	\end{table}

@@

	%%\begin{figure}[htb]
	%%
	%%\includegraphics[scale=0.05]{Figs/dynamiSelectionSchedules}
	%%\caption{Task set and schedules where orders are precomputed based on different heuristics.}
	%%\label{fig:dynamicSelectionExample}
	%%\end{figure}

@@

	In this class of strategy, we precompute the order of tasks based on some criterion and then follow this ordering as much as possible. But when the communication resource is idle because the memory requirement of the next task is too high, then we select a task  with a dynamic strategy. After a task is dynamically selected, we update the remaining order without this task. This class of strategy takes advantage of static information in the form of precomputed order and dynamic corrections to minimize the idle time due to memory constraint.
	%%%%alg_name="optimal_order_infinite_memory_largest_communication \_ \allowbreak task \_ \allowbreak respects \_ \allowbreak memory \_ \allowbreak restriction";
	%%%%string alg_name="optimal_order_infinite_memory_smallest_communication \_ \allowbreak task \_ \allowbreak respects \_ \allowbreak memory \_ \allowbreak restriction";
	%%%%string alg_name="optimal_order_infinite_memory_maximum_accelerated \_ \allowbreak task \_ \allowbreak respects \_ \allowbreak memory \_ \allowbreak restriction";

@@

			\begin{tabular}{|c|c|c|c|}
				\hline
				%%				\multirow{2}{*}{Task} & Comm Time & \multirow{2}{*}{Comp Time}\\ 
				%%				&=Comm Volume& \\ \hline
				Task & Comm Volume & Comm Time & Comp Time\\ \hline 
				A & 4 & 4 &  1\\ \hline
				B & 2 & 2 & 6\\ \hline
				C & 8 & 8 & 8\\ \hline
				D & 5 & 5 & 4\\ \hline
				E & 3 & 3 & 2\\ \hline
			\end{tabular}
			\caption{\label{tab:staticOrderDynamicCorrectionsExample} A task set for static order dynamic corrections schedules.}
		\end{center}
	\end{table}

@@

		\begin{table}[htb]
			%%	\scriptsize
			\begin{tabular}{|c|p{6.5cm}|}
				\hline
				\textbf{Heuristic} & \textbf{\hspace{2cm}Favorable Situation} \\ \hline
				$OOSIM$ & Memory capacity is not a restriction (\textcolor{green}{Optimal}) \\ \hline
				$IOCMS$ & Memory capacity is not a restriction and tasks are compute intensive (\textcolor{green}{Optimal}) \\ \hline
				$DOCPS$ & Memory capacity is not a restriction and tasks are communication intensive (\textcolor{green}{Optimal}) \\ \hline
				$IOCCS$ & Moderate memory capacity and most tasks are highly compute intensive \\ \hline
				$DOCCS$ & Moderate memory capacity and most tasks are highly communication intensive \\ \hline
				$LCMR$ & Limited memory capacity and significant percentage of tasks with large communication timings are compute intensive\\ \hline
				$SCMR$ & Limited memory capacity and significant percentage of tasks with small communication timings are compute intensive\\ \hline
				$MAMR$ & Limited memory capacity and significant percentage of tasks of both types\\ \hline
				$OOLCMR$ & Moderate memory capacity and significant percentage of tasks are communication intensive\\ \hline
				$OOSCMR$ & Moderate memory capacity and significant percentage of tasks are compute intensive \\ \hline
				$OOMAMR$ & Moderate memory capacity and significant percentage of highly compute and communication intensive tasks \\ \hline
			\end{tabular}\caption{~\label{tab:heuristicsAndFavorableScenarios}Heuristics and their favorable scenarios.}
		\end{table}

@@

		\section{Experimental Results}
		\label{sec:exp}

		%%	\todo[inline]{Machine information and our model}
		\subsection{Experimental Settings}

		We consider a machine called Cascade~\cite{Cascade}, available at PNNL, for our experiments. We obtain traces by running two molecular chemistry applications, double precision version of Hartree-Fock (HF) and Coupled Cluster Single Double (CCSD) of NWChem\todo{cite or url} package on 10 nodes of this machine. Each node is composed of 16 Intel Xeon E5-2670 cores. NWChem takes advantages of a Partitioned Global Address Space Programming Model called Global Arrays (GA)~\cite{GlobalArray} to use shared-memory programming APIs on distributed memory computers. GA dedicates one core of each node to handle other cores, hence we can view a node as being composed of 15 computational cores. We use 150 processes for each application and obtain 150 trace files. We run CCSD with Uracil molecules input and HF with SiOSi molecules (for Uracil molecules, HF has a much smaller workload, each processor executes only around 20 tasks, that is why we chose SiOSi input for HF execution). Each process executes around 300-800 tasks. Our data transfer model is quite simple and we consider that all data transfers between the local memory of each process and the GA memory take the same route. Modeling of different routes of data transfers for the same source-destination pair, bandwidth sharing for different source-destination pairs and network congestion is more challenging and part of our future work. This simple model is enough to provide insight to the application developers (or runtime system) about the ordering of data transfers for the same source-destination pair so as to maximize communication-computation overlap. Our model is easily adaptable to any source-destination pair when there is one fixed route between source and destination (such as between CPU and GPU, one copy engine to transfer data from CPU (resp. GPU) to GPU (resp. CPU) ).
		\todo[inline]{More about tiling}
		Both applications mainly perform two types of computations, tensor transpose and tensor contraction. HF expects to specify a tile size and we set it to 100, so that each core can be used efficiently. CCSD automatically determines tile sizes at different program points based on the input molecules. Hence, HF operates on almost homogeneous tiles while CCSD uses more heterogeneous tiles.

@@

		%%		\begin{figure}[htb]
		%%			%\begin{tabular}{c}
		%%			\subfloat[HF workloads\label{fig:hfProperties}]{%
		%%				\includegraphics[width=.45\linewidth]{./results/plots/application_properties_hf.pdf}
		%%			}%\\
		%%			\subfloat[CCSD workloads\label{fig:ccsdProperties}]{%
		%%				\includegraphics[width=.45\linewidth]{./results/plots/application_properties_ccsd.pdf}
		%%			}%\\
		%%			%\end{tabular}
		%%			\caption{HF and CCSD tasks characteristics.}
		%%			\label{fig:ApplicationProperties}
		%%		\end{figure}

@@

		%% 
		%%\begin{figure}[htb]
		%%	\centering
		%%	\begin{subfigure}{.5\textwidth}
		%%		\centering
		%%		\includegraphics[width=.95\linewidth]{../ExperimentalResults/application_properties_hf.pdf}
		%%		\caption{HF workloads}
		%%		\label{fig:hfProperties}
		%%	\end{subfigure}%
		%%	\begin{subfigure}{.5\textwidth}
		%%		\centering
		%%		\includegraphics[width=.95\linewidth]{../ExperimentalResults/application_properties_ccsd.pdf}
		%%		\caption{CCSD workloads}
		%%		\label{fig:ccsdProperties}
		%%	\end{subfigure}
		%%	\caption{HF and CCSD tasks characteristics.}
		%%	\label{fig:ApplicationProperties}
		%%\end{figure}

@@

		%%		\todo[inline]{increase font size in all plots, may be free scale-y}
		%%	\todo[inline]{SKumar: decide metrics: ratio to optimal, percentage overlap, 100 - percentage non overlap}

@@

		Figures ~\ref{fig:ratio_to_optimal_hf} and ~\ref{fig:ratio_to_optimal_ccsd} show the distribution of the performance of each heuristic for all considered memory capacities, where plots are categorized by memory capacities. For each memory capacity and each heuristic, the box on the plot displays the median, first and last quartile, and the whiskers indicate minimum and maximum values, with outliers are shown by black dots.
		\subsection{HF Performance}	
		%%	\begin{figure*}[htb]
		%%		\includegraphics[scale=0.25]{./results/plots/ratio_to_optimal_selected_hf.pdf}
		%%		\caption{Comparison of different heuristics for HF.}
		%%		\label{fig:ratio_to_optimal_hf}
		%%	\end{figure*}
		\begin{figure*}[htb]
			\includegraphics[scale=0.5]{./all-binpack/ratio_to_optimal_selected_hf.pdf}
			\caption{Comparison of different heuristics for HF.\todo[inline]{LED: Wouldn't it be easier to read if facet labels were expressed as ratios of $m_c$ ($1$, $1.125$, $1.25$, ...) instead of absolute values? Skumar: It'll provide a nice view but $m_c$ values of both experiments are quite different. In that case, we have to explicitly mention $m_c$ values.}}
			\label{fig:ratio_to_optimal_hf}
		\end{figure*}

@@

		\subsection{CCSD Performance}

		%%	\begin{figure*}[htb]
		%%		\includegraphics[scale=0.25]{./results/plots/ratio_to_optimal_selected_ccsd.pdf}
		%%		\caption{Comparison of different heuristics for CCSD.}
		%%		\label{fig:ratio_to_optimal_ccsd}
		%%	\end{figure*}

		\begin{figure*}[htb]
			\includegraphics[scale=0.5]{./all-binpack/ratio_to_optimal_selected_ccsd.pdf}
			\caption{Comparison of different heuristics for CCSD.}
			\label{fig:ratio_to_optimal_ccsd}
		\end{figure*}

@@

		\subsection{Scheduling in Batches}
		In most applications, the runtime scheduler may only observe a limited batch of independent tasks. Therefore we organize tasks of each trace file in batches of 100 (the last batch may have less than 100 tasks). We apply each heuristic on each group independently. Figure~\ref{fig:best_variants_batch} shows the performance of the best variants of each category for both applications. The plots exhibit behavior similar to Figures~\ref{fig:ratio_to_optimal_best_hf} and~\ref{fig:ratio_to_optimal_best_ccsd}: static order with dynamic correction variants attain maximum communication-computation overlap and outperform other heuristics.
		%%		\todo[inline]{SKumar:combine both plots}

		%%		
		%%		\begin{figure*}[htb]
		%%			\subfloat[Best variants of HF.\label{fig:hf_best_variants_batch}]{%
		%%				\includegraphics[width=.45\linewidth]{./results/plots/inverse_ratio_to_optimal_hf_batch-best.pdf}
		%%			}%\\
		%%			\subfloat[Best variants of CCSD.\label{fig:ccsd_best_variants_batch}]{%
		%%				\includegraphics[width=.45\linewidth]{./results/plots/inverse_ratio_to_optimal_ccsd_batch-best.pdf}
		%%			}%\\
		%%			\caption{Best variants of all categories where heuristics are applied in the batches of 100 tasks.}
		%%			\label{fig:best_variants_batch}
		%%		\end{figure*}

>>>>

	\section{Data Transfer Problem Formulation}
	\label{sec:theoreticalProof}

@@

@@

	\begin{table}
		\begin{center}
			\begin{tabular}{|c|c|c|c|}
				\hline
				Task & Comm Volume & Comm Time & Comp Time\\ \hline 
				A & 0 & 0 & 5\\ \hline
				B & 4 & 4 & 3\\ \hline
				C & 1 & 1 & 6\\ \hline
				D & 3 & 3 & 7\\ \hline
				E & 6 & 6 & 0.5\\ \hline
				F & 7 & 7 & 0.5\\ \hline
			\end{tabular}
		\end{center}
		\caption{Example instance where ordering on both resources
			has to be different.}
		\label{table:different.order}
	\end{table}

@@

@@

	\subsection{Dynamic Selection}
	Dynamic strategies are based on the following principle: when the communication resource is idle, a task is chosen based on a ranking criterion which differ depending on the heuristic, among those which fit in memory. For example, if the selection criteria is to choose a highly compute intensive task, then we compute the ratio of compute time and communication time for all tasks, and we select the task which fit in the currently available memory and whose ratio is maximum (ties are broken arbitrarily). If no task satisfies the selection criteria thee we leave the resource idle at that point and proceed to the next event point. We also ensure that the order on both resources is the same, by ordering tasks on the computing resource accordingly.

@@

	%% We select a set of tasks which induce minimum idle time on compute resource and satisfy available memory requirement. Then a task is chosen for processing from this set based on the following heuristics.
	\begin{enumerate}[label=\roman*)]
		\item \textit{largest communication task respects memory restriction} ($LCMR$): a task with the largest communication time is chosen. 
		\item \textit{smallest communication task respects memory restriction} ($SCMR$): a task with the smallest communication time is chosen.
		\item \textit{maximum accelerated task respects memory restriction} ($MAMR$): a task with the maximum ratio of compute time to communication time is chosen.
	\end{enumerate}
	\begin{table}[htb]
		\begin{center}

			\begin{tabular}{|c|c|c|c|}
				\hline
				Task & Comm Volume & Comm Time & Comp Time\\ \hline 
				A & 3 & 3 & 2\\ \hline
				B & 1 & 1 &  6\\ \hline
				C & 4 & 4 & 6\\ \hline
				D & 5 & 5 & 1\\ \hline
			\end{tabular}
			\caption{\label{tab:dynamicSelectionExample} A task set for dynamic schedules.}
		\end{center}
	\end{table}

@@

@@

	In this class of strategy, we precompute the order of tasks based on some criterion and then follow this ordering as much as possible. But when the communication resource is idle because the memory requirement of the next task is too high, then we select a task  with a dynamic strategy. After a task is dynamically selected, we update the remaining order without this task. This class of strategy takes advantage of static information in the form of precomputed order and dynamic corrections to minimize the idle time due to memory constraint.

@@

			\begin{tabular}{|c|c|c|c|}
				\hline
				Task & Comm Volume & Comm Time & Comp Time\\ \hline 
				A & 4 & 4 &  1\\ \hline
				B & 2 & 2 & 6\\ \hline
				C & 8 & 8 & 8\\ \hline
				D & 5 & 5 & 4\\ \hline
				E & 3 & 3 & 2\\ \hline
			\end{tabular}
			\caption{\label{tab:staticOrderDynamicCorrectionsExample} A task set for static order dynamic corrections schedules.}
		\end{center}
	\end{table}

@@

		\begin{table}[htb]
			\begin{tabular}{|c|p{6.5cm}|}
				\hline
				\textbf{Heuristic} & \textbf{\hspace{2cm}Favorable Situation} \\ \hline
				$OOSIM$ & Memory capacity is not a restriction (\textcolor{green}{Optimal}) \\ \hline
				$IOCMS$ & Memory capacity is not a restriction and tasks are compute intensive (\textcolor{green}{Optimal}) \\ \hline
				$DOCPS$ & Memory capacity is not a restriction and tasks are communication intensive (\textcolor{green}{Optimal}) \\ \hline
				$IOCCS$ & Moderate memory capacity and most tasks are highly compute intensive \\ \hline
				$DOCCS$ & Moderate memory capacity and most tasks are highly communication intensive \\ \hline
				$LCMR$ & Limited memory capacity and significant percentage of tasks with large communication timings are compute intensive\\ \hline
				$SCMR$ & Limited memory capacity and significant percentage of tasks with small communication timings are compute intensive\\ \hline
				$MAMR$ & Limited memory capacity and significant percentage of tasks of both types\\ \hline
				$OOLCMR$ & Moderate memory capacity and significant percentage of tasks are communication intensive\\ \hline
				$OOSCMR$ & Moderate memory capacity and significant percentage of tasks are compute intensive \\ \hline
				$OOMAMR$ & Moderate memory capacity and significant percentage of highly compute and communication intensive tasks \\ \hline
			\end{tabular}\caption{~\label{tab:heuristicsAndFavorableScenarios}Heuristics and their favorable scenarios.}
		\end{table}

@@

		\section{Experimental Results}
		\label{sec:exp}
		\subsection{Experimental Settings}

		We consider a machine called Cascade~\cite{Cascade}, available at PNNL, for our experiments. We obtain traces by running two molecular chemistry applications, double precision version of Hartree-Fock (HF) and Coupled Cluster Single Double (CCSD) of NWChem~\cite{NWChem} package on 10 nodes of this machine. Each node is composed of 16 Intel Xeon E5-2670 cores. NWChem takes advantages of a Partitioned Global Address Space Programming Model called Global Arrays (GA)~\cite{GlobalArray} to use shared-memory programming APIs on distributed memory computers. GA dedicates one core of each node to handle other cores, hence we can view a node as being composed of 15 computational cores. We use 150 processes for each application and obtain 150 trace files. We run CCSD with Uracil molecules input and HF with SiOSi molecules (for Uracil molecules, HF has a much smaller workload, each processor executes only around 20 tasks, that is why we chose SiOSi input for HF execution). Each process executes around 300-800 tasks. Our data transfer model is quite simple and we consider that all data transfers between the local memory of each process and the GA memory take the same route. Modeling of different routes of data transfers for the same source-destination pair, bandwidth sharing for different source-destination pairs and network congestion is more challenging and part of our future work. This simple model is enough to provide insight to the application developers (or runtime system) about the ordering of data transfers for the same source-destination pair so as to maximize communication-computation overlap. Our model is easily adaptable to any source-destination pair when there is one fixed route between source and destination (such as between CPU and GPU, one copy engine to transfer data from CPU (resp. GPU) to GPU (resp. CPU) ).
		\todo[inline]{More about tiling}
		Both applications mainly perform two types of computations, tensor transpose and tensor contraction. HF expects to specify a tile size and we set it to 100, so that each core can be used efficiently. CCSD automatically determines tile sizes at different program points based on the input molecules. Hence, HF operates on almost homogeneous tiles while CCSD uses more heterogeneous tiles.

@@

@@

@@

@@

		Figures ~\ref{fig:ratio_to_optimal_hf} and ~\ref{fig:ratio_to_optimal_ccsd} show the distribution of the performance of each heuristic for all considered memory capacities, where plots are categorized by memory capacities. For each memory capacity and each heuristic, the box on the plot displays the median, first and last quartile, and the whiskers indicate minimum and maximum values, with outliers are shown by black dots.
		\subsection{HF Performance}	
		\begin{figure*}[htb]
			\includegraphics[scale=0.5]{./all-binpack/ratio_to_optimal_selected_hf.pdf}
			\caption{Comparison of different heuristics for HF.\todo[inline]{LED: Wouldn't it be easier to read if facet labels were expressed as ratios of $m_c$ ($1$, $1.125$, $1.25$, ...) instead of absolute values? Skumar: It'll provide a nice view but $m_c$ values of both experiments are quite different. In that case, we have to explicitly mention $m_c$ values.}}
			\label{fig:ratio_to_optimal_hf}
		\end{figure*}

@@

		\subsection{CCSD Performance}
		\begin{figure*}[htb]
			\includegraphics[scale=0.5]{./all-binpack/ratio_to_optimal_selected_ccsd.pdf}
			\caption{Comparison of different heuristics for CCSD.}
			\label{fig:ratio_to_optimal_ccsd}
		\end{figure*}

@@

		\subsection{Scheduling in Batches}
		In most applications, the runtime scheduler may only observe a limited batch of independent tasks. Therefore we organize tasks of each trace file in batches of 100 (the last batch may have less than 100 tasks). We apply each heuristic on each group independently. Figure~\ref{fig:best_variants_batch} shows the performance of the best variants of each category for both applications. The plots exhibit behavior similar to Figures~\ref{fig:ratio_to_optimal_best_hf} and~\ref{fig:ratio_to_optimal_best_ccsd}: static order with dynamic correction variants attain maximum communication-computation overlap and outperform other heuristics.
